---
title: "Reproduction of eca estimates"
author: "Edvin Fuglebakk"
date: "5/23/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../compare_results_multiple_stox_runs.R")
eca_results <- list()
eca_results$kolmule2017 <- "../kolmule/eca_2017/KOLMULE2017.fit.2017_allgears_totalarea_season1234.caa.txt"
eca_results$makrell2018 <- "../makrell/eca_2018/MAKRELL2018.fit.2018_allgears_totalarea_season1234.caa.txt"
eca_results$sei.nssk.2018 <- "../sei_nssk/eca_2018/SEI2018_ns_v00.fit.2018_allgears_totalarea_season1234.caa.txt"
eca_results$hyse.nssk.2018 <- "../hyse_nssk/eca_2018/HYSE2018_ns_v00.fit.2018_allgears_27fire_season1234.caa.txt"
eca_results$torsk.nssk.2018 <- "../torsk_nssk/eca_2018/TORSK2018_ns_v00.fit.2018_allgears_IV_season1234.caa.txt"
stox_results <- list()
stox_results$kolmule2017 <- "../kolmule/stox_2017/results"
stox_results$makrell2018 <- "../makrell/stox_2018/results"
stox_results$sei.nssk.2018 <- "../sei_nssk/stox_2018/results/"
stox_results$hyse.nssk.2018 <- "../hyse_nssk/stox_2018/results/"
stox_results$torsk.nssk.2018 <- "../torsk_nssk/stox_2018/results/"

```

## Background
ECA 3.x are implementations of The ECA model (Hirst et al. 2004, 2005 and 2012) developed by the Norwegian Computing Centre for IMR. ECA 3.x has been used for catch at age estimations for several species. The Norwegian Computing Centre has developed an new implementation of ECA delivered to IMR and the REDUS project as an R-package which I will refer to as RECA. As part oif the REDUS project an data APIs and a user interface for configuring and running Reca has been implemented in StoX. I will refer to estimates produced via this interfaces as StoX-Reca estimates. The StoX-Reca integration aims to achieve several goals:

* Provide an implementation of ECA that is acceptable for user that have decided on using ECA for catch at age estimates.
* Provide a user interface familiar to StoX users.
* Separate data handling and estimation, so that filtering and transformation are exposed to users.
* Document all model parameters and data versions for purposes of recomputability.

The first StoX-Reca version has been delivered and has been tested during the spring of 2019.

### Design decisions
Some decisions was made during the StoX-Reca implementation that excludes certain model and date configurations possible with the ECA 3.x. 

* It has not yet been implemented a procedure for adjusting landings data with logbook records. Such procedure was supported via separate scripts for ECA 3.x. Adding this to the bookeeping necessary for versioning data is an extra burden, and the value added to the estimates has not been analysed.
* Grouping of covariate values is not supported in StoX-Reca, rather it allows for redefinig variables to different resolutions. This is more flexible and can in most cases mimic grouping.
* Configuring covariates differently for the underlying ECA models ("length given age" and "weight given length") is not supported in StoX-Reca, as this requires grouping covariate values.
* Not all filters that are supported by ECA 3.x are implemented in StoX-Reca. Notably filtering for outliers in the age-length scatter. We have decided we want to flag these as data errors in stead, to be filtered out explicitly or cleaned up in the database.

## Usability testing
Configuring estimates for reproduction has revealed some usability issues that should be considered for further StoX-Reca development.

* The landings format readable by StoX is still only provided via API calls. A user interface in datasetexplorer should be considered.
* At times I found the landings data to be outdated, not reflecting recent corrections from FDIR. A procedure for regularly updating the landings API should be defined.
* Grouping of areas are common in the configurations I have attempted to reproduce. Currently this require generation of new stratafiles and neighbourmatrices in StoX, which is cumbersome for users. A user interface for redefining spatial covariate and neighbour-matrices should be considered. Alternativelty scripts for generating regrouping of common strata systems could be provided
* Some configurations use grouping of seasons. The way to do this is not documented in stox. 
* Configuration of reports. It is common to configure ECA to decompose estimates on other covariates then those configured for the model (e.g. configure model for F.dir-areas, and reports for ICES areas). Stox-processes to support that should be considered.
* ECA 3.x supports filtering data based on whether there are aged fish in hauls. StoX only supports this via the field sampletype, which is sensitive to punching errors and to whether all ages have been read at the time of analysis.
* Estimates for specific stocks often require filtering data by area. If only positions are available, this is best done after stratum is assigned. But this requires filtering on variables added to the data in StoX, which I think is not supported.
* At the end of the prepareRECA process, data checks are performed and reports on potential data issues are compiled. Currently prepareRECA halts on some of these data issues, which precludes the reports from being compiled. Extracting data issue-reporting to a process run prior to prepareRECA should be considered.
* Reports on data issues compiles report for common covariates, without checking if these are configured. Consider if this can be avoided.

## Reproduction of ECA 3.x results
As part of assuring the quality of the StoX-Reca implementations, I have attempted to configure StoX-Reca projects that mimic ECA 3.x estimates that has been used for advice. We have not had protocols for complete documentation and model configuration for estimates. Indeed that is a central task for the ongoing incorporation of commercial catch estimates into StoX. I therefore can not identify all data and all model configurations from previous ECA 3.x runs. In cases where I have ECA .param files I have near complete description of model configuration, but exactly identifying the data seems impossible. I have in these cases had to tinker with data selection to demonstrate that a reasonable data selection that approximately reproduces previous results can be found. This tinkering is similar to what will have to be done to generate time-series. Unlest explicitly stated otherwise, this consist only of filtering on mission types and sample types. Selection of specific stations have only been done in order to match a complete listing of serial numbers from the ECA 3.x run, when such are available. ECA estimates are however dependent not only on model configuration and data, but also on random parameter exploration in the Monte Carlo simulations, and therefore also on the extent of Monte Carlo sampling. Luckily, I have documentation for the extent of simulations (number of samples and burnin-iterations) for all projects. Difference in estimated parameters that are smaller than random differences in parameter exploration does not affect valid interpretations of the results and I will therefore compare ECA 3.x estimates to multiple StoX-Reca estimates, generated with different random seed. Even considering this spread of estimates, there are several obstacles to exact estimation. As mentioned above some features of ECA 3.x has been deliberately left out of the StoX-Reca framework, and minor bugs have been identified and corrected in ECA throughout the integration process. I think we shall therefore be content with reproduction of estimates that seem consistent with historical results up to differences that would only have minor consequences for advice. I will refer to reproductions where mean and standard deviation for age groups are within the variability expected between several ECA-runs to be *strictly reproduced*. Reproductions that I think have difference of only minor consequence, I will refer to as *sufficiently reproduced*. 

Using the criteria outlned above, I aim to reproduce the following estimates:

* blue whiting 2017: *Strictly reproduced*
* mackerel 2018: *Inconclusive*
* Norwegian Spring Spawning Herring 2017: *software error*
* NSSK saithe 2018: *Sufficiently reproduced*
* NSSK haddock 2018: *Inconclusive*
* NSSK cod 2018: *Sufficiently reproduced*

I will just give brief descriptions of configurations below. Consult StoX project files for details.

### Reproduction of blue whiting 2017
ECA 3.x estimate from 2017. Model was configured and run by Sondre Aanes and results provided by Are Salthaug. ECA 3.x .param files was not kept. Based on Sondres notes, I have configured the model with only temporal (season) and platform (boat) as covariates, and verified that station selection (serial numbers) was identical between the two estimates. Quarter 2-4 was collapsed to one value. I used samples from HRF (High-seas reference fleet) and commercial samples, and ran 1500 burnin iterations and 500 sampling iterations, with the log-linear growth model.

* Total catch used in ECA 3.x estimate was 399 kt
* Total catch used in StoX-Reca estimae was 399 kt

Both mean estimates and standard deviations from ECA 3.x are consistent with the range of ten estimates from StoX-Reca. I consider this a strict reproduction of the estimates for the 2018 advice.

```{r kolmule means, fig.cap={"Estimates for mean catch in millions per age group of blue whiting in 2017."}, echo=FALSE}

comppar("kolmule2017", "mean")
```

```{r kolmule sd, fig.cap={"standard deviation of marginal posterior distribution of catch in millions per age group of blue whiting in 2017."}, echo=FALSE}

comppar("kolmule2017", "sd")
```

### Reproduction of mackerel 2018
ECA 3.x estimate from 2018. Model was configured and run by Sondre Aanes who provided results for reproduction analysis. ECA 3.x .param files was not kept, but a detailed log of data corrections done by ECA 3.x was provided. However, 1964 individuals were listed as removed for 'unkown reason'. I could not find a way to identify these. Total number of individuals in StoX-Reca run was ~4K, total number in ECA 3.x sample overview was ~2K.

* Total catch used in ECA 3.x estimate was 187 kt
* Total catch used in StoX-Reca estimae was 187 kt

In order to get a close reproduction of standard deviations, I ran the estimate without platform as a random-effect covariate (boat effect). This is not what was done in the estimate that Sondre sent. It is in my opinion not easy to explain how the differences in specimen selection should compensate the use of boat effect so exactly, and I'd like to re-run the ECA 3.x estimate to make sure that boat effect was actually used. 

~~Most age groups are well within the intervals reported with the previous estimate, age group 1 just barely outside. I consider this a sufficient reproduction of the estimates for the 2019 advice.~~

```{r makrell means, fig.cap={"Estimates for mean catch in millions per age group of mackerel in 2018."}, echo=FALSE}

comppar("makrell2018", "mean")
```

```{r makrell sd, fig.cap={"standard deviation of marginal posterior distribution of catch in millions per age group of mackerel in 2018."}, echo=FALSE}

comppar("makrell2018", "sd")
```

```{r makrell comparison, fig.cap={"StoX-Reca mean estimates on ECA 3.x 90% credible intervals"}, echo=FALSE}
compareConfBands("makrell2018")
```


### Reproduction of Norwegian Spring Spawning herring 2017
ECA 3.x estimate from 2017. Model was configured and run by Sondre Aanes and results provided by Are Salthaug. ECA 3.x .param files was not kept. Based on Sondres notes, I have filtered an exact serialnumber list, grouped Gillnet with Trawl, grouped seasons 1 with two and 3 with 4 and run 1000 burnin iterations and 500 sampling iterations with non-linear growth model.

* Total catch used in ECA 3.x estimate was 389 kt
* Total catch used in StoX-Reca estimate was 389 kt

Running this configuration triggered a bug in the RECA library. This has been reported to the Norwegian Computing Centre and Estimates will be revisited when a revised library is delivered.

### Reproduction of saithe NSSK 2018
ECA 3.x estimates were prepared by Åge Fotland who provided results for reproduction analysis. This included ECA 3.x .param files. The ECA 3.x estimates used logbook-adjusted landings and had weight-length model configured with different geargroupings than the age-length model, neither of which is currently supported by StoX-Reca. Parameters extracted from ECA 3.x .param files suggest the model was configured with season and gear as fixed effects, but I did not have enough aged samples in the data compiled for reproduction to configure it like that. I configured the temporal covariate as a random effect in stead.

* Total catch used in ECA 3.x estimate was 40 kt
* Total catch used in StoX-Reca estimate was 40 kt 


Allt age groups are well within the intervals reported with the previous estimate. I consider this a sufficient reproduction of the estimates for the 2019 advice.


```{r sei means, fig.cap={"Estimates for mean catch in millions per age group of saithe in the North Sea and Skagerak 2018."}, echo=FALSE}

comppar("sei.nssk.2018", "mean")
```

```{r sei sd, fig.cap={"standard deviation of marginal posterior distribution of catch in millions per age group of saithe in the North Sea and Skagerak 2018."}, echo=FALSE}

comppar("sei.nssk.2018", "sd")
```

```{r sei comparison, fig.cap={"StoX-Reca mean estimates on ECA 3.x 90% credible intervals"}, echo=FALSE}
compareConfBands("sei.nssk.2018")
```


### Reproduction of haddock NSSK 2018
ECA 3.x estimates were prepared by Åge Fotland who provided results for reproduction analysis. This included ECA 3.x .param files. The ECA 3.x estimates used logbook-adjusted landings, which is not currently supported by StoX-Reca. Parameters extracted from ECA 3.x .param files suggest the model was configured with season and gear as fixed effects, but I did not have enough aged samples in the data compiled for reproduction to configure it like that. I configured the temporal covariate as a random effect in stead. The spatial covariate had only one value, such covariates are implicitly dropped by ECA 3.x, so I configured the StoX-Reca esitmate without the spatial covariate.

* Total catch used in ECA 3.x estimate was 1399 t
* Total catch used in StoX-Reca estimate was 1433 t

This discrepancy is probably due to the use of logbook-adjusted landings in ECA 3.x

The estimation shows a severe discrepancy for age 1. Inspection of data used reveals that there is likely less data in the StoX-Reca run (~50 stations), and that both estimates have a very low number of fish samples (aged specimens) supporting this age group (<5). The serialnumbers used was documented by Åge Fotland. Pulling these from the database reveals that most are missing mandataroy fields for estimating number in haul (catchsampleweight or lengthsamplecount). A few of these had catchcount listed, which are not used ny StoX-Reca. I have consulted Sondre Aanes, which can confirm that ECA 3.x are tolerant to missing number in haul estimates for samples which are not partioned before sampling (delprøve). This likely explains the difference in age group 1. Missing haul weights are generally fixable (from logbooks or landings), and it might not be necessary for StoX-Reca to support estimating when these are not present. In order to support missing sample weights, we may consider checking if the data
requirements can be relaxed. In this regard I have asked NR to clarify the documentation.

```{r hyse means, fig.cap={"Estimates for mean catch in millions per age group of haddock in the North Sea and Skagerak 2018."}, echo=FALSE}

comppar("hyse.nssk.2018", "mean")
```
```{r hyse sd, fig.cap={"standard deviation of marginal posterior distribution of catch in millions per age group of haddock in the North Sea and Skagerak 2018."}, echo=FALSE}

comppar("hyse.nssk.2018", "sd")
```

```{r hyse comparison, fig.cap={"StoX-Reca mean estimates on ECA 3.x 90% credible intervals"}, echo=FALSE}
compareConfBands("hyse.nssk.2018")
```

### Reproduction of cod NSSK 2018
ECA 3.x estimates were prepared by Åge Fotland who provided results for reproduction analysis. This included ECA 3.x .param files. The ECA 3.x estimates used logbook-adjusted landings, which is not currently supported by StoX-Reca. Results were only provided decomposed on ICES area III (Skagerak) and IV (North Sea). I recall from conversation with Åge that coastal regions are excluded from these estimates, so I filtered landings by kysthav-code. I included coastal samples, still. 

#### ICES area IV
Parameters extracted from ECA 3.x .param files suggest the model was configured with season and gear as fixed effects, but I did not have enough samples with weight in the data compiled for reproduction to configure it like that. I configured the temporal covariate as a random effect in stead. I run the analysis for ICES area III and IV separately, rather than decoposing the report. In that case, the spatial covariate had only one value, such covariates are implicitly dropped by ECA 3.x, so I configured the StoX-Reca esitmate without the spatial covariate.

* Total catch used in ECA 3.x estimate was 5133 t
* Total catch used in StoX-Reca estimate was 5197 t

This discrepancy is probably due to the use of logbook-adjusted landings in ECA 3.x

Most age groups are well within the intervals reported with the previous estimate, age group 1 just barely outside. I consider this a sufficient reproduction of the estimates for the 2019 advice.


```{r torsk means, fig.cap={"Estimates for mean catch in millions per age group of cod in the North Sea 2018."}, echo=FALSE}

comppar("torsk.nssk.2018", "mean")
```

```{r torsk sd, fig.cap={"standard deviation of marginal posterior distribution of catch in millions per age group of cod in ICES Sub Area 4."}, echo=FALSE}

comppar("torsk.nssk.2018", "sd")
```

```{r torsk comparison, fig.cap={"StoX-Reca mean estimates on ECA 3.x 90% credible intervals"}, echo=FALSE}
compareConfBands("torsk.nssk.2018")
```

#### Effect of updating landings
I plan to re-run the NSSK estimates with later updates to the 2018 landings data, so that the effect of landings corrections can be compared the discrepancies between logbook-corrected landings and "raw"-landings that is particularly evident in the cod estimates. I will follow up on this when NMD has implemented protocols for updating landings (est. oct 2019).